\section{Conclusion}
\label{sec:conclusion}

\subsection{Cross-Model Comparison}

Table~\ref{tab:final_comparison} summarises the performance of all models on
the full test set using MAE as the primary metric. Each model is evaluated on the
same held-out test period; the improvement column is computed relative to the
raw linear regression baseline (MAE~$= 0.638$).

\begin{table}[H]
  \centering
  \caption{Cross-model comparison (MAE on test set, standardised target).
    Lower is better.}
  \label{tab:final_comparison}
  \begin{tabular}{lrr}
    \toprule
    Model & MAE (full) & vs.\ linear baseline \\
    \midrule
    Linear Regression (raw lags)                    & 0.638  & --- \\
    Linear Regression (feature-engineered)          & 0.598  & $-6\,\%$ \\
    GARCH(1,1)$^\dagger$                            & 0.00583 & --- \\
    FNN (ReLU, $L=3$, SGD $\mu=0.95$)              & 0.579  & $-9\,\%$ \\
    RNN ($T=20$, $H=64$, Adam)                     & 0.502  & $-21\,\%$ \\
    \bottomrule
    \multicolumn{3}{l}{\footnotesize
      $^\dagger$ GARCH is estimated on unscaled returns; its MAE is not
      directly comparable to the other four models.}
  \end{tabular}
\end{table}

\noindent Every neural network model outperforms the linear baseline, and the
gap widens as architectural complexity increases. The RNN, which explicitly
models the temporal ordering of lags, achieves the largest gain: a
\textbf{21\,\% reduction in MAE} relative to the raw linear regression, or
\textbf{13\,\%} relative to the best FNN configuration.

\subsection{Key Findings}

Three themes emerge consistently across all architectures.

\paragraph{Non-linearity matters, but is not enough.}
Moving from linear regression to an FNN reduces MAE from~$0.638$ to~$0.579$
($-9\,\%$). The ReLU activation allows the network to learn the approximate
symmetry $f(x) \approx f(-x)$ required for volatility prediction --- a
representation that is fundamentally unavailable to a linear model with signed
inputs. Feature engineering (squared lags) partially compensates for this
limitation without introducing learned non-linearity, bringing the linear
baseline to~$0.598$. However, neither linear nor FNN models exploit the
\emph{ordering} of lags: their performance is largely insensitive to input
permutation.

\paragraph{Temporal structure is a genuine source of signal.}
The RNN's 13\,\% improvement over the FNN demonstrates that the sequential
ordering of past returns carries predictive information beyond their set-level
statistics. The model learns to assign greater sensitivity to recent lags and
to accumulate context across the input window --- a capability unavailable to a
flat feature-vector architecture. Yet the gain is partially undermined by
vanishing gradients: output sensitivity collapses within approximately one
recurrence step, limiting the effective memory despite a 20-lag input window.

\paragraph{Regime blindness is the dominant unsolved challenge.}
Across all architectures, crisis-period MAE is \textbf{2.5--4$\times$ higher}
than calm-period MAE. During stress episodes (dot-com crash, 2008 GFC, COVID
shock), every model systematically under-predicts volatility because the
training distribution offers few examples of extreme moves. The ARCH signal
that makes forecasting tractable during calm periods breaks down precisely when
forecasting accuracy matters most --- during crises. This gap is not closed by
switching from linear to non-linear models, nor by introducing temporal memory:
it reflects a fundamental data scarcity problem that cannot be resolved by
architectural choices alone.

\subsection{Limitations and Future Work}

The results presented in this report point to several natural extensions.

\paragraph{Gated recurrent architectures.}
The vanilla RNN's vanishing-gradient problem is directly addressed by the Long
Short-Term Memory (LSTM) and Gated Recurrent Unit (GRU) architectures, which
introduce multiplicative gating mechanisms that decouple the gradient highway
from the output projection. Based on the grid search results (Section~\ref{sec:rnn}),
the ceiling imposed by architecture rather than hyperparameters suggests that a
gated model is the single most promising next step.

\paragraph{Regime-aware modelling.}
All models in this study treat calm and crisis observations as draws from the
same distribution. A natural extension is to condition the model on an estimated
regime state --- either via an explicit regime-switching layer, a Hidden Markov
Model that labels observations before training, or a threshold variable (e.g.\
the VIX index) provided as an additional input feature.

\paragraph{Multivariate inputs.}
The forecasting task here relies exclusively on lagged log-returns of a single
asset. Incorporating cross-asset information (implied volatility indices, credit
spreads, macro factors) could provide leading indicators of regime transitions
that are absent from the univariate input.

\paragraph{Attention mechanisms.}
Transformer-based architectures replace recurrence with self-attention, allowing
the model to attend directly to any lag in the input window without the gradient
attenuation that limits RNNs. This could address the long-range dependency
problem identified in the vanishing-gradient analysis while remaining fully
parallelisable during training.
